---
title: "Covid Data"
output: html_document

---
## Analysis of the covid dataset and analyse the relationship between the occurance-rate and population.

Covid data has been analysed several ways, and the simplest kind of analysis is to join it with the deaths
and perform various visualizaitons and analysis baseed on that. This has been done before and several times.
I thought it would be interesting to download the population dataset from cencus.gov and see if we can identify
counties that do better or worse in December of 2021.

## Biases

### Bias in data collection.
Several biases may occur at the point of data collection. There was a lot of misinformation going around during 
covid and it is possible that some counties that show low case count may have under-reported.

### Personal biases
I believe in the scientific method and that most people/gov officials/doctors acted in good faith during the pandemic. This attitude, I hope helps me overcome any personal bias in my analysis. The analysis is not based on race,age or gender and therefore I don't think there is any scope for personal bias affecting the result 


## Data Source :

"time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv"

### Lets download the data
```{r}
library(tidyverse)

cov_us <- read_csv('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv')

#cov_global <- read_csv('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv')

#cov_us_deaths <- read_csv('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv')

#cov_global_deaths <- read_csv('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv')

head(cov_us)
#head(cov_us_deaths)

```
### Convert Covid US data long table and fiter by date = 2020-12-01 
```{r}
cov_us_2020 <- cov_us %>% mutate(Combined_Key=paste0(Province_State," ", Admin2, ' County'))  %>%
    select(-c('UID','iso2','iso3','code3','FIPS','Admin2', 'Lat', 'Long_')) %>% 
    pivot_longer(cols=-c('Province_State', 'Country_Region', 'Combined_Key'), names_to='dates', values_to='cases') %>%
    mutate(dates=as.Date(dates, '%m/%d/%y')) %>% 
    mutate(year_month=as.Date(format(dates, '%Y-%m-01'))) %>% 
    filter(dates=='2020-12-01')
cov_us_2020$year <- '2020'

```
### Get population data at county level from cencus.gov
Download the data, clean and rearrange
```{r}
df_pop <- read_csv('https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv')
df_pop <- df_pop %>% select(c('STNAME','CTYNAME','ESTIMATESBASE2020','POPESTIMATE2020','POPESTIMATE2021','POPESTIMATE2022','POPESTIMATE2023'))
df_pop$combined_key <- paste0(df_pop$STNAME, " ", df_pop$CTYNAME)
df_pop_year <- df_pop %>% rename('2020'=POPESTIMATE2020, '2021'=POPESTIMATE2021, '2022'=POPESTIMATE2022, '2023'=POPESTIMATE2023) %>% pivot_longer(cols=-c('STNAME', 'CTYNAME', 'combined_key'), names_to = 'date', values_to = 'population')
```

### Join the two datasets

#### What the counties/cities with the highest occurance as a % of the population?
```{r}
df_us_occ <- left_join(cov_us_2020, df_pop_year, by=join_by(Combined_Key==combined_key, year==date))
df_us_occ$perc_occ <- df_us_occ$cases*100/df_us_occ$population
head(df_us_occ %>% arrange(desc(perc_occ), as.integer(year_month))) %>% select(Combined_Key, population, perc_occ)
```
In December 2020, Colorado Crowlery County had the higest no of cases as a % of the population. But thats a relatively smaller county. Does the picture change for a county with population > 100000?

```{r}
head(df_us_occ %>% filter(population> 100000 & year=='2020') %>% arrange(desc(perc_occ), as.integer(year_month))) %>% select(Combined_Key, population, cases, perc_occ)

```

It is now the Potter County in Texas with a rate of 11.75 % in Dec 2020!

#### Lets plot the cumulative cases over tie for Texas Potter County
```{r}
cov_piv <- cov_us %>% mutate(Combined_Key=paste0(Province_State," ", Admin2, ' County'))  %>%
    select(-c('UID','iso2','iso3','code3','FIPS','Admin2', 'Lat', 'Long_')) %>% 
    pivot_longer(cols=-c('Province_State', 'Country_Region', 'Combined_Key'), names_to='dates', values_to='cases') %>%
    mutate(dates=as.Date(dates, '%m/%d/%y'))
  
  
cov_piv %>% filter(Combined_Key=='Texas Potter County') %>% 
  arrange(cases) %>% select(Combined_Key, dates, cases) %>% ggplot(aes(x=dates, y=cases )) + ggtitle("Cummulative cases for Texas Potter County") + geom_line()
```

What is the relationship between population and occurrence? We'll also add a line of best fit
``` {r}
df1 <- df_us_occ %>% filter(population>10000 & population<1000000) %>% arrange(population) %>% select(Combined_Key, population, cases, perc_occ) 

fit <- lm(perc_occ ~ population, data=df1)
df1$pred <- predict(fit, df1)
df1 %>% ggplot(aes(x=population, y=perc_occ)) + geom_point() + geom_line(aes(y=pred))
```

Its obvious that there is a relationship between the population of a county and the number of cases. Howver we see that
that for counties with smaller populaiton there is a huge spread, with some counties doing much worse than the others.  

#### What are the counties with low population and high occurance ?

```{r}
head(df_us_occ %>% select(Combined_Key, perc_occ, population) %>% filter(population < 10000 & perc_occ > 10) %>% arrange(desc(perc_occ)))
```


#### What are the counties with low population and low occurance?

```{r}
head(df_us_occ %>% select(Combined_Key, perc_occ, population) %>% filter(population < 10000 & perc_occ <1) %>% arrange(perc_occ))
```

So in December 2020 when the pandemic was at its peak, there were countines with 0 case load. Exactly how many ?

```{r}
df_us_occ %>% select(Combined_Key, perc_occ, population) %>% filter(perc_occ==0) %>% arrange(desc(population))
```
 Around 25 countly, mostly in Utah, with Weber county having 1/4 million population have 0 cases! Are these rural isolated communities? It would be awesome to put these on map and see if a pattern emerges.
 
 *P.S*: The mapping packages are difficult to install on machine and I am leaving that as an exercise for next time.
 
 
 